# Basic syntax

```{r, include=FALSE}
if (!"SciViews:R" %in% search())
  source("R/SciViews.R", attach(NULL, name = "SciViews:R"))
```

R provides a rich working environment with lot of tools. In order to get everything organised and easier to manage, most features are provided through **functions**. Installable ensembles of function, data sets, documentation and examples are called R **packages**. At a higher organisational level, SciViews::R provides **sets** (a series of related functions) and **units** (a remix of the content of one or more packages).

## Functions

function

:   A function in computer language performs one task in a separate unit of code. A function accepts zero or more **arguments** that passes variables and parameters and usually returns one object as a result. Most functions have no **side-effect**, that is, they modifies nothing in your computing environment. Some do, however, like printing a result, drawing a chart, changing a global option, writing in a file, ...

functional language

:   A computing language that relies on functions to organize its code by applying and composing them.

A call to a function consist in its symbol, followed by parentheses, and arguments provided inside these parentheses.

The *definition* of the function can be see with `args(<function_name>)` :

![](figures/functions_arguments1_drawio){fig-align="left"}

The *call* of the function in R code can specify arguments *by position* or *by their name*:

![](figures/functions_arguments2_drawio)

For instance, the variance of a numeric vector is computed by the `var()` function like this:

```{r}
x <- c(1, 4, 3.2, 6.3)
var(x)
```

Here, we provided one argument to `var()` : `x` , the numeric vector. The `var()` function has more arguments. One get a tip in RStudio, or VSCode, or one can rely on `args()` to see these arguments, or display the help page with `?var` to get even more information.

```{r}
args(var)
```

Also, running just `var` without parentheses prints the code of the function.

**Recurrent pattern:** a function uses a data frame and a formula. Often, the argument matching the formula is the first one. However, it is more logical to indicate data first. It can be done with a named argument:

![](figures/functions_data_formula_drawio)

When using the pipe operator `|>` with R \>= 4.2.0, one can use `data = _` . Here is a linear model using `lm()` that way:

```{r}
mtcars |>
  lm(data = _, mpg ~ hp + cyl)
```

The **Call:** in the output shows arguments are correctly interpreted, including the formula as first argument of `lm()`.

generic function

:   A generic function, or just "generic" is a function that runs different code depending on the **class** of one or more of its arguments (but usually, of the first argument).

method

:   A function that implements the code of a generic for one particular class. There may be also a **default** method that applies if no specific method is found for the generic

dispatching

:   The selection of the right method given the context (call to a generic with argument(s) of given class(es)

Example: the average of a vector of numeric values is calculated by the `mean()` function. But `mean()` is generic. So, using `args()` , or printing its code does not give much information:

```{r}
args(mean)
mean
```

Here, we see that besides its first argument named `x` , there is a variable set of additional arguments materialized by the three dots `…` In fact the actual arguments depends on the class of `x` . The code is limited to `UseMethod("mean")`, which indicate to dispatch on "mean" for the first argument.

-   `r fn_list("methods()", "base::methods()", "object")` lists all known methods for a generic

```{r}
methods(mean)
```

We see that there is **default** method, as well as a series of others. Those with an asterisk (\*) are hidden and not directly accessible.

-   `r fn_list("get_s3_methods()", "base::getS3methods()", "object")` gets the code for a given S3 method (S3 is a type of R objects, see the objects chapter)

```{r}
get_s3_method('mean', 'default')
```

## Packages

## Sets

## Units

## Branching

Describe here if, while, repeat, for.

## Syntactic sugar

Describe here {, (, pipe, dot list, invisible...

Next section will present basic math operators.
